---
title: "MAS31004 - individual project"
author: "yuanyuan wei"
date: "2024-11-13"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r setup2, include=FALSE, echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE}
set.seed(1234)
```

The analysis employs R packages like dplyr, ggplot2, and tidyr for data manipulation and visualization.

The dataset was cleaned to remove missing values, ensuring a reliable analysis of total medals, weighted scores, and economic indicators.
```{r}
#import the packages 
library(dplyr)

library(ggplot2)

library(tidyr)

library(carData)

library(car)  # For VIF calculation
```


```{r}
#load the data
setwd("~/Desktop")
data <- read.csv("MedalsStatsData-220251110.csv")

#rename the columns
library(dplyr)

data <- data %>%

  rename(

    total_medals = total_medals,

    weighted_score = weighted_score,

    medals_per_capita = medals_per_capita,

    population = population,

    GDP = GDP..in.millions.,

    GDP_per_capita = GDP_per_capita

  )

print(colnames(data))
```

```{r}
#check the missing values 
sapply(data[c("total_medals", "weighted_score", "medals_per_capita", "population", "GDP", "GDP_per_capita")], function(x) sum(is.na(x)))

#clean the data
data_clean <- na.omit(data[c("total_medals", "weighted_score", "medals_per_capita", "population", "GDP", "GDP_per_capita")])
```

```{r}
descriptive_stats <- data_clean %>%

  summarise(

    total_medals_mean = mean(total_medals, na.rm = TRUE),

    total_medals_median = median(total_medals, na.rm = TRUE),

    total_medals_range = paste(min(total_medals, na.rm = TRUE), max(total_medals, na.rm = TRUE), sep = " to "),

    weighted_score_mean = mean(weighted_score, na.rm = TRUE),

    weighted_score_median = median(weighted_score, na.rm = TRUE),

    weighted_score_range = paste(min(weighted_score, na.rm = TRUE), max(weighted_score, na.rm = TRUE), sep = " to "),

    medals_per_capita_mean = mean(medals_per_capita, na.rm = TRUE),

    medals_per_capita_median = median(medals_per_capita, na.rm = TRUE),

    medals_per_capita_range = paste(min(medals_per_capita, na.rm = TRUE), max(medals_per_capita, na.rm = TRUE), sep = " to ")

  )

print(descriptive_stats)
```

# Correlation

The correlation matrix reveals important relationships between our variables:

-   Strong Positive Correlations:
    -   Total Medals and Weighted Score: r = 0.9965
    -   Population and GDP: r = 0.8104
    -   Total Medals and GDP: r = 0.7695
    -   Weighted Score and GDP: r = 0.7846
-   Moderate Positive Correlations:
    -   Total Medals and Population: r = 0.3925
    -   Total Medals and GDP per Capita: r = 0.2692
-   Weak Positive Correlations:
    -   Medals per Capita and Population: r = -0.1211
    -   Medals per Capita and GDP: r = -0.1149
-   Negligible Correlation:
    -   Medals per Capita and GDP per Capita: r = -0.0076
    
Moderate to weak correlations suggest that larger or wealthier countries do not always perform better on a per capita basis.
```{r}
correlation_matrix <- cor(data_clean)

print(correlation_matrix)
```

# Visual Analysis

```{r}
pairs(data_clean, pch = 21, bg = c("pink"))
```

# Regression

```{r}
model_total_medals <- lm(total_medals ~ population + GDP + GDP_per_capita, data = data_clean)

model_weighted_score <- lm(weighted_score ~ population + GDP + GDP_per_capita, data = data)

model_medals_per_capita <- lm(medals_per_capita ~ GDP_per_capita, data = data)

summary(model_total_medals)

summary(model_weighted_score)

summary(model_medals_per_capita)

#Check for multicollinearity

vif(model_total_medals)

vif(model_weighted_score)
```

## graphs

```{r}
#total medals log scale graph 
ggplot(data, aes(x = GDP, y = total_medals)) +

  geom_point(color = "blue", alpha = 0.6, size = 2) +  # Increase point size

  geom_smooth(method = "lm", color = "red", se = TRUE) +  # Add confidence interval

  scale_x_log10() +  # Log scale for GDP

  labs(title = "Total Medals vs. GDP (Log Scale)",

       x = "GDP (in millions, log scale)",

       y = "Total Medals") +

  theme_minimal(base_size = 15) +  # Increase base size for readability

  theme(plot.title = element_text(hjust = 0.5),  # Center title

        panel.grid.major = element_line(color = "grey80"),  # Light grid lines

        panel.grid.minor = element_blank(),  # Remove minor grid lines

        axis.text = element_text(size = 12),  # Axis text size

        axis.title = element_text(size = 14))  # Axis title size
```

```{r}
# Create the Weighted Score vs. GDP graph with enhancements

ggplot(data, aes(x = GDP, y = weighted_score)) +

  geom_point(color = "green", alpha = 0.6, size = 2) +  # Increase point size

  geom_smooth(method = "lm", color = "orange", se = TRUE) +  # Add confidence interval

  scale_x_log10() +  # Log scale for GDP

  labs(title = "Weighted Score vs. GDP (Log Scale)",

       x = "GDP (in millions, log scale)",

       y = "Weighted Score") +

  theme_minimal(base_size = 15) +  # Increase base size for readability

  theme(plot.title = element_text(hjust = 0.5),  # Center title

        panel.grid.major = element_line(color = "grey80"),  # Light grid lines

        panel.grid.minor = element_blank(),  # Remove minor grid lines

        axis.text = element_text(size = 12),  # Axis text size

        axis.title = element_text(size = 14))  # Axis title size
```

```{r}
#medals per capita vs GDP per capita
ggplot(data, aes(x = GDP_per_capita, y = medals_per_capita)) +

  geom_point(color = "purple", alpha = 0.6) +

  geom_smooth(method = "lm", color = "magenta") +

  labs(title = "Medals Per Capita vs. GDP Per Capita",

       x = "GDP Per Capita",

       y = "Medals Per Capita") +

  theme_minimal()
```

# Regional Analysis

```{r}
library(dplyr)
regional_stats <- data %>%

  group_by(region) %>%

  summarise(

    avg_total_medals = mean(total_medals, na.rm = TRUE),

    avg_weighted_score = mean(weighted_score, na.rm = TRUE),

    avg_medals_per_capita = mean(medals_per_capita, na.rm = TRUE)

  )

print(regional_stats)
```



```{r}
#ANOVA for regional differences
anova_total_medals <- aov(total_medals ~ region, data = data)
summary(anova_total_medals)

avg_weighted_score <- aov(weighted_score ~ region, data = data)
summary(avg_weighted_score)

avg_medals_per_capita <- aov(medals_per_capita ~ region, data = data)
summary(avg_medals_per_capita)
```

# Visualisation

```{r}
# Visualization: Box plots for regional comparison

library(ggplot2)

ggplot(data, aes(x = region, y = total_medals)) +

  geom_boxplot() +

  theme_minimal() +

  labs(title = "Distribution of Total Medals by Region", x = "Region", y = "Total Medals") +

  theme(

    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)  # Adjust size and angle of x-axis text

  )
```


```{r}
# Visualization: Box plots for regional comparison of Weighted Scores 
library(ggplot2)

ggplot(data, aes(x = region, y = weighted_score)) +

  geom_boxplot() +

  theme_minimal() +

  labs(title = "Distribution of Weighted Score by Region", x = "Region", y = "Weighted Score") +

  theme(

    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)  # Adjust size and angle of x-axis text

  )
```

```{r}
# Visualization: Box plots for regional comparison of Medals per Capita
library(ggplot2)

ggplot(data, aes(x = region, y = medals_per_capita)) +

  geom_boxplot() +

  theme_minimal() +

  labs(title = "Distribution of Medals per Capita by Region", x = "Region", y = "Medals per Capita") +

  theme(

    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)  # Adjust size and angle of x-axis text

  )
```

# efficiency analysis

We calculated an efficiency index to identify countries that perform exceptionally well relative to their economic resources:
```{r}
#Efficiency Analysis

data <- data %>%

  mutate(efficiency_index = weighted_score / (GDP / 1e9))

top_efficient_countries <- data %>%

  arrange(desc(efficiency_index)) %>%

  select(country, efficiency_index) %>%

  head(10)

print(top_efficient_countries)
```

```{r}
rmarkdown::render("MedalsStatsScript-220251110.Rmd", output_format="pdf_document")
```








